---
title: "Tuning Multiple Similar Parameters: Ensuring `autoplot` Uniqueness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning Multiple Similar Parameters: Ensuring `autoplot` Uniqueness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("keras") &&
    requireNamespace("keras3", quietly = TRUE) &&
    requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("tune", quietly = TRUE) &&
    requireNamespace("dials", quietly = TRUE) &&
    requireNamespace("parsnip", quietly = TRUE) &&
    requireNamespace("workflows", quietly = TRUE) &&
    requireNamespace("recipes", quietly = TRUE) &&
    requireNamespace("rsample", quietly = TRUE)
)
```

## Introduction

When using `kerasnip` to define and tune Keras models within the `tidymodels` framework, you might encounter situations where you want to tune multiple parameters that, by default, map to the same underlying `dials` parameter type. A common example is tuning the number of units in multiple `layer_dense` blocks within the same model.

While `kerasnip` intelligently maps these parameters (e.g., `dense1_units` and `dense2_units` both map to `dials::hidden_units()`), this can lead to ambiguity when visualizing tuning results with `ggplot2::autoplot()`. Without a way to distinguish between these otherwise identical parameter types, `autoplot()` may produce errors or misleading plots.

This vignette demonstrates how to explicitly provide unique identifiers to your tuned parameters, ensuring `autoplot()` can correctly visualize the results for each distinct parameter.

## The Problem (Implicit)

Consider a model with two dense layers, each with a `units` parameter. If you were to define them for tuning without unique `id`s, `autoplot()` would encounter an issue because it cannot distinguish between the two parameters.

For example, if you were to run `ggplot2::autoplot(tune_res)` without unique `id`s, you might encounter an error similar to this:

```{r}
#> Error in `dplyr::rename()`:
#> ! Names must be unique.
#> ✖ These names are duplicated:
#>   * "# Hidden Units" at locations 1 and 2.
```

This error clearly indicates that `autoplot()` is trying to rename columns for plotting, but it finds duplicate names like "# Hidden Units" because both `dense1_units` and `dense2_units` are generically identified as `hidden_units` by `dials` without further distinction. This makes it impossible for `autoplot()` to differentiate their tuning results.

## The Solution: Using Unique `id`s with `tune()`

The solution is to provide a unique `id` argument to the `tune()` function for each parameter you wish to distinguish.

Let's define a simple sequential Keras model with two dense layers:

```{r model_definition}
library(kerasnip)
library(keras3)
library(parsnip)
library(dials)
library(workflows)
library(recipes)
library(rsample)
library(tune)
library(ggplot2)

# Define a spec with multiple hidden unit parameters
model_name <- "autoplot_unique_spec"
# Clean up the spec if it already exists from a previous run
if (exists(model_name, mode = "function")) {
  suppressMessages(remove_keras_spec(model_name))
}

input_block <- function(model, input_shape) {
  keras3::keras_model_sequential(input_shape = input_shape)
}

dense_block <- function(model, units = 10) {
  model |> keras3::layer_dense(units = units)
}

output_block <- function(model, num_classes) {
  model |>
    keras3::layer_dense(units = num_classes, activation = "softmax")
}

create_keras_sequential_spec(
  model_name = model_name,
  layer_blocks = list(
    input = input_block,
    dense1 = dense_block,
    dense2 = dense_block,
    output = output_block
  ),
  mode = "classification"
)

# Now, create the model specification and assign unique IDs for tuning
tune_spec <- autoplot_unique_spec(
  dense1_units = tune(id = "dense_layer_one_units"),
  dense2_units = tune(id = "dense_layer_two_units")
) |>
  set_engine("keras")

print(tune_spec)
```

Notice how `dense1_units` and `dense2_units` are both passed to `tune()`, but each is given a distinct `id`. This `id` acts as a label that `autoplot()` can use to differentiate the parameters.

### Setting up the Tuning Workflow

Next, we'll set up a `tidymodels` workflow, define the parameter ranges, and create a tuning grid.

```{r tuning_setup}
# Set up workflow and tuning grid
rec <- recipes::recipe(Species ~ ., data = iris)
tune_wf <- workflows::workflow(rec, tune_spec)

params <- tune::extract_parameter_set_dials(tune_wf)

# Update the parameter ranges using kerasnip::hidden_units
# The `id`s provided in tune() are automatically detected and used here.
params <- params |>
  update(
    dense1_units = hidden_units(range = c(4L, 8L)),
    dense2_units = hidden_units(range = c(4L, 8L))
  )

grid <- dials::grid_regular(params, levels = 2)
control <- tune::control_grid(save_pred = FALSE, verbose = FALSE)

print(grid)
```

### Running the Tuning Process

Now, we run `tune::tune_grid` to perform the actual tuning. For demonstration purposes, we'll use a small number of resamples and a simple dataset.

```{r run_tuning}
# Run tuning
tune_res <- tune::tune_grid(
  tune_wf,
  resamples = rsample::vfold_cv(iris, v = 2),
  grid = grid,
  control = control
)

print(tune_res)
```

### Visualizing Results with `autoplot()`

With the tuning complete, we can now use `ggplot2::autoplot()` to visualize the results. Because we provided unique `id`s, `autoplot()` can correctly generate separate plots for each tuned parameter.

```{r autoplot_results, fig.width=7, fig.height=5}
# Assert that autoplot works without error
ggplot2::autoplot(tune_res)
```

As you can see, `autoplot()` successfully generates a plot showing the performance across the different values for `dense_layer_one_units` and `dense_layer_two_units` independently.

## Why Unique `id`s are Necessary

Internally, `kerasnip` maps arguments like `units` from your `layer_blocks` functions to appropriate `dials` parameter objects (e.g., `dials::hidden_units()`). When multiple such arguments exist, they all point to the *same type* of `dials` parameter.

The `id` argument in `tune()` serves as a unique identifier that `tune` and `ggplot2::autoplot()` use to distinguish between different instances of these parameter types. Without it, `autoplot()` would see multiple parameters of type `hidden_units` and wouldn't know how to plot them separately, leading to errors or combining them incorrectly.

## Best Practices

*   **Always use unique `id`s:** When tuning multiple parameters that are conceptually similar (e.g., `units` in different layers, `rate` in different dropout layers), always provide a descriptive and unique `id` to the `tune()` function.
*   **Descriptive `id`s:** Choose `id`s that clearly indicate which part of the model the parameter belongs to (e.g., `dense_layer_one_units`, `conv_filter_size`). This improves readability and understanding of your tuning results.

By following this practice, you ensure that your `kerasnip` models are robustly tunable and that their results can be clearly visualized using the `tidymodels` ecosystem.

```{r cleanup, include=FALSE}
suppressMessages(remove_keras_spec(model_name))
```
