---
title: "Prediction Intervals with Conformal Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction Intervals with Conformal Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("keras")
)
options(keras.fit_verbose = 0)
set.seed(123)
```

## Why neural networks do not produce native prediction intervals

Linear models derive prediction intervals analytically from the covariance structure of their parameter estimates. A neural network has no equivalent: it computes a single deterministic output for each input, with no internal distributional model to derive intervals from.

This is not a limitation specific to `kerasnip`. The entire tidymodels ecosystem reflects the same reality, the `parsnip` package does not register `conf_int` or `pred_int` prediction types for any neural network engine, including `brulee`, the official tidymodels neural network package.

The supported and recommended path for obtaining prediction intervals from neural networks in tidymodels is **conformal inference**, provided by the `probably` package. Conformal inference is:

- **Distribution-free**: it makes no assumptions about the shape of the outcome distribution.
- **Model-agnostic**: it treats the model as a black box; only `fit()` and `predict()` calls are required.
- **Guaranteed to have correct coverage**: under the assumption that training and test data are exchangeable (i.e. identically and independently distributed), the intervals contain the true outcome with at least the requested probability.

The `probably` package provides three conformal methods, described in the sections below.

## Setup

```{r setup}
library(kerasnip)
library(tidymodels)
library(probably)
```

## Defining a kerasnip regression model

All three methods work with any fitted `workflows::workflow()` object, so the setup is identical to any other kerasnip workflow.

```{r define-spec}
input_block <- function(model, input_shape) {
  keras3::keras_model_sequential(input_shape = input_shape)
}
dense_block <- function(model, units = 32) {
  model |>
    keras3::layer_dense(units = units, activation = "relu")
}
output_block <- function(model) {
  model |> keras3::layer_dense(units = 1)
}

create_keras_sequential_spec(
  model_name = "conf_mlp",
  layer_blocks = list(
    input  = input_block,
    dense  = dense_block,
    output = output_block
  ),
  mode = "regression"
)

spec <- conf_mlp(dense_units = 32, fit_epochs = 30) |>
  set_engine("keras")

data <- modeldata::ames |> select(
  Sale_Price,
  Gr_Liv_Area,
  Year_Built,
  Garage_Area,
  Total_Bsmt_SF
)
rec <- recipe(Sale_Price ~ ., data = data) |>
  step_normalize(all_numeric_predictors())
wflow <- workflow(rec, spec)
```

---

## Method 1: Split conformal inference (`int_conformal_split`)

This is the simplest and fastest method. The workflow is:

1. Fit the model on a **training set**.
2. Pass a separate **calibration set** to `int_conformal_split()`. The function runs `predict()` once on the calibration set and records the residuals.
3. Call `predict()` on the resulting object to obtain intervals for new data.

The model is never re-fitted. The calibration residuals act as a reference distribution for deciding how wide to make the intervals.

```{r split-conformal}
set.seed(1)
split <- initial_split(data, prop = 0.75)
train_dat <- training(split)
cal_dat <- testing(split)

# Fit on the training set
fit_obj <- fit(wflow, data = train_dat)

# Build the conformal object from the calibration set
conformal_split <- int_conformal_split(fit_obj, cal_data = cal_dat)
conformal_split

# Predict intervals for new observations
new_obs <- cal_dat[1:6, ]
predict(conformal_split, new_data = new_obs, level = 0.90)
```

**When to use this**: when training cost is non-trivial and you can afford to
set aside a calibration set. This is the recommended method for kerasnip
models in most practical situations.

---

## Method 2: Cross-validation conformal inference (`int_conformal_cv`)

This method uses **cross-validation** to generate out-of-fold predictions, then pools those residuals to calibrate the intervals. The workflow is:

1. Define cross-validation folds.
2. Run `tune::fit_resamples()` on the workflow with `save_pred = TRUE` and `extract = function(x) x` so that both predictions and the fitted model objects are retained for each fold.
3. Pass the resulting `resample_results` object to `int_conformal_cv()`. It reads the held-out predictions and uses them to calibrate the intervals.
4. Call `predict()` on the resulting object to obtain intervals for new data.

The final model is fitted once more on the full training data inside `int_conformal_cv()`.

```{r cv-conformal}
set.seed(1)
folds <- vfold_cv(data, v = 5)

# Fit the workflow on each fold, retaining predictions and model objects
fitted_folds <- tune::fit_resamples(
  wflow,
  folds,
  control = tune::control_resamples(
    save_pred = TRUE,
    extract = function(x) x
  )
)

# Build the conformal object from the resampling results
conformal_cv <- int_conformal_cv(fitted_folds)
conformal_cv

predict(conformal_cv, new_data = data[1:6, ], level = 0.90)
```

**When to use this**: when you do not want to reserve a separate calibration set and are already running cross-validation to evaluate your model. The cross-validation residuals tend to produce slightly narrower intervals than the split method because they use more of the data. For kerasnip models, note that this refits the model once per fold, which multiplies training time by the number of folds.

---

## Method 3: Full conformal inference (`int_conformal_full`)

Full conformal inference is the most principled of the three methods in terms of statistical guarantees, but it is the most expensive computationally.

The algorithm works observation by observation. For each new test point, it appends that observation to the training set with a **candidate outcome value**, refits the model from scratch on the augmented dataset, and checks whether the residual for the candidate value is consistent with the training residuals. By sweeping over a grid of candidate values, it determines the interval of values that the algorithm would not reject.

### Why `int_conformal_full` is unreliable without `fit_seed`

For a kerasnip model, every refit goes through a full build–compile–train cycle starting from **random weight initialisation**. This creates a fundamental incompatibility with `probably`'s bounds-search algorithm, which assumes that as the candidate outcome sweeps across the trial grid, the nonconformity score for the trial row changes **monotonically**. That assumption holds for deterministic models, linear regression, random forests, gradient boosting, where the same data always produces the same model. It does not hold for a neural network re-trained from a different random state on every refit.

When the scores happen to be non-monotone, `probably` cannot find the interval boundaries and returns `NA`. **Critically, this does not happen on every run.** The outcome depends on the data, the architecture, and the random state of the session at the time of the call. Results may look valid on one run and produce `NA` on the next, with no way to know in advance which will occur. A result that appears correct cannot be trusted to be reproducible.

kerasnip detects when it is being called inside `int_conformal_full`'s refit loop and emits a warning on every internal refit when `fit_seed` is absent. The warning is the **reliable signal**, it fires unconditionally regardless of whether the current run happens to succeed:

```{r full-conformal-broken, eval = FALSE}
# This code pattern will trigger a kerasnip warning on every internal refit.
# The intervals may or may not be NA depending on the session's random state.
fit_small <- fit(wflow, data = data_small)

conformal_full_broken <- int_conformal_full(
  fit_small,
  train_data = data_small,
  control = control_conformal_full(
    method = "grid",
    trial_points = 20
  )
)

# Warning: int_conformal_full() refits the model from scratch for every trial
# value. Without `fit_seed`, each refit starts from a different random
# initialisation, so nonconformity scores across the trial grid may be
# non-monotone. When that happens, probably cannot find the interval boundaries
# and returns NA — but the problem does not occur on every run, so results that
# look valid may not be reproducible or trustworthy.
# Fix: add `fit_seed = <integer>` to your model spec ...
predict(conformal_full_broken, new_data = new_obs_small, level = 0.90)
```

If you see this warning, treat the output as unreliable regardless of whether the intervals contain numbers or `NA`.

### The fix: `fit_seed`

Setting `fit_seed` in the model spec calls `keras3::set_random_seed()` at the start of every internal refit, making each training run deterministic and restoring the monotonicity the bounds-search requires. The warning will not fire and the intervals will be reproducible across runs:

```{r full-conformal-fixed}
# Use a small subset to keep runtime reasonable in this vignette.
# int_conformal_full refits the model (n_new_obs x n_grid_points) times —
# here 3 x 20 = 60 complete training runs on top of the initial fit.
data_small    <- data[1:100, ]
new_obs_small <- data[101:103, ]

# fit_seed = 42L ensures every internal refit uses the same random state.
# The value is arbitrary; what matters is that it is fixed and consistent.
spec_seeded <- conf_mlp(
  dense_units = 32,
  fit_epochs  = 30,
  fit_seed    = 42L
) |>
  set_engine("keras")

wflow_seeded <- workflow(rec, spec_seeded)
fit_seeded   <- fit(wflow_seeded, data = data_small)

conformal_full <- int_conformal_full(
  fit_seeded,
  train_data = data_small,
  control = control_conformal_full(
    method = "grid",
    trial_points = 20
  )
)
conformal_full

predict(conformal_full, new_data = new_obs_small, level = 0.90)
```

### When to use `int_conformal_full`

Prefer `int_conformal_split` or `int_conformal_cv` for kerasnip models, both calibrate a single fixed model and carry none of the above caveats. Use `int_conformal_full` only when the stronger coverage guarantee is required, the training cost is acceptable, and `fit_seed` is set.

---

## Comparison of the three methods

| Method | Model refits | Coverage guarantee | Recommended for kerasnip |
|---|---|---|---|
| `int_conformal_split` | 1 (on training data only) | Marginal | ✅ Yes, lowest cost |
| `int_conformal_cv` | 1 per fold + 1 final | Marginal | ✅ Yes, if already cross-validating |
| `int_conformal_full` | *(n_test × n_grid)* | Marginal | ⚠️ Only with `fit_seed`, unreliable otherwise |

All three methods provide **marginal coverage**: across many test observations, at least the requested fraction will have their true outcome within the interval. None of them provide **conditional coverage** (intervals that are exactly correct for every specific input value), but in practice they are well-calibrated for most regression problems.

---

## Cleanup

```{r cleanup}
remove_keras_spec("conf_mlp")
```
