---
title: "Prediction Intervals with Conformal Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction Intervals with Conformal Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("keras")
)
options(keras.fit_verbose = 0)
set.seed(123)
```

## Why neural networks do not produce native prediction intervals

Linear models derive prediction intervals analytically from the covariance structure of their parameter estimates. A neural network has no equivalent: it computes a single deterministic output for each input, with no internal distributional model to derive intervals from.

This is not a limitation specific to `kerasnip`. The entire tidymodels ecosystem reflects the same reality, the `parsnip` package does not register `conf_int` or `pred_int` prediction types for any neural network engine, including `brulee`, the official tidymodels neural network package.

The supported and recommended path for obtaining prediction intervals from neural networks in tidymodels is **conformal inference**, provided by the `probably` package. Conformal inference is:

- **Distribution-free**, it makes no assumptions about the shape of the outcome distribution.
- **Model-agnostic**, it treats the model as a black box; only `fit()` and `predict()` calls are required.
- **Guaranteed to have correct coverage**, under the assumption that training and test data are exchangeable (i.e. identically and independently distributed), the intervals contain the true outcome with at least the requested probability.

The `probably` package provides three conformal methods, described in the sections below.

## Setup

```{r setup}
library(kerasnip)
library(tidymodels)
library(probably)
```

## Defining a kerasnip regression model

All three methods work with any fitted `workflows::workflow()` object, so the setup is identical to any other kerasnip workflow.

```{r define-spec}
input_block <- function(model, input_shape) {
  keras3::keras_model_sequential(input_shape = input_shape)
}
dense_block <- function(model, units = 32) {
  model |>
    keras3::layer_dense(units = units, activation = "relu")
}
output_block <- function(model) {
  model |> keras3::layer_dense(units = 1)
}

create_keras_sequential_spec(
  model_name = "conf_mlp",
  layer_blocks = list(
    input  = input_block,
    dense  = dense_block,
    output = output_block
  ),
  mode = "regression"
)

spec <- conf_mlp(dense_units = 32, fit_epochs = 30) |>
  set_engine("keras")

data <- modeldata::ames |> select(
  Sale_Price,
  Gr_Liv_Area,
  Year_Built,
  Garage_Area,
  Total_Bsmt_SF
)
rec <- recipe(Sale_Price ~ ., data = data) |>
  step_normalize(all_numeric_predictors())
wflow <- workflow(rec, spec)
```

---

## Method 1: Split conformal inference (`int_conformal_split`)

This is the simplest and fastest method. The workflow is:

1. Fit the model on a **training set**.
2. Pass a separate **calibration set** to `int_conformal_split()`. The function runs `predict()` once on the calibration set and records the residuals.
3. Call `predict()` on the resulting object to obtain intervals for new data.

The model is never re-fitted. The calibration residuals act as a reference distribution for deciding how wide to make the intervals.

```{r split-conformal}
set.seed(1)
split <- initial_split(data, prop = 0.75)
train_dat <- training(split)
cal_dat <- testing(split)

# Fit on the training set
fit_obj <- fit(wflow, data = train_dat)

# Build the conformal object from the calibration set
conformal_split <- int_conformal_split(fit_obj, cal_data = cal_dat)
conformal_split

# Predict intervals for new observations
new_obs <- cal_dat[1:6, ]
predict(conformal_split, new_data = new_obs, level = 0.90)
```

**When to use this**: when training cost is non-trivial and you can afford to
set aside a calibration set. This is the recommended method for kerasnip
models in most practical situations.

---

## Method 2: Cross-validation conformal inference (`int_conformal_cv`)

This method uses **cross-validation** to generate out-of-fold predictions, then pools those residuals to calibrate the intervals. The workflow is:

1. Define cross-validation folds.
2. Run `tune::fit_resamples()` on the workflow with `save_pred = TRUE` and `extract = function(x) x` so that both predictions and the fitted model objects are retained for each fold.
3. Pass the resulting `resample_results` object to `int_conformal_cv()`. It reads the held-out predictions and uses them to calibrate the intervals.
4. Call `predict()` on the resulting object to obtain intervals for new data.

The final model is fitted once more on the full training data inside `int_conformal_cv()`.

```{r cv-conformal}
set.seed(1)
folds <- vfold_cv(data, v = 5)

# Fit the workflow on each fold, retaining predictions and model objects
fitted_folds <- tune::fit_resamples(
  wflow,
  folds,
  control = tune::control_resamples(
    save_pred = TRUE,
    extract = function(x) x
  )
)

# Build the conformal object from the resampling results
conformal_cv <- int_conformal_cv(fitted_folds)
conformal_cv

predict(conformal_cv, new_data = data[1:6, ], level = 0.90)
```

**When to use this**: when you do not want to reserve a separate calibration set and are already running cross-validation to evaluate your model. The cross-validation residuals tend to produce slightly narrower intervals than the split method because they use more of the data. For kerasnip models, note that this refits the model once per fold, which multiplies training time by the number of folds.

---

## Method 3: Full conformal inference (`int_conformal_full`)

Full conformal inference is the most principled of the three methods in terms of statistical guarantees, but it is the most expensive computationally.

The algorithm works observation by observation. For each new test point, it appends that observation to the training set with a **candidate outcome value**, refits the model from scratch on the augmented dataset, and checks whether the residual for the candidate value is consistent with the training residuals. By sweeping over a grid of candidate values, it determines the interval of values that the algorithm would not reject.

For a kerasnip model, every refit call goes through the full build–compile–train cycle: the Keras architecture is reconstructed from the layer blocks, compiled, and trained from random weight initialisation. This means the total number of training runs is:

> *(number of new test observations)* × *(number of candidate grid points)*

For the example below with 6 new observations and a 20-point grid, that is 120 complete model training runs. Even with 30 epochs on a small dataset this takes meaningful time. In practice, `int_conformal_split` or `int_conformal_cv` are the recommended choices for kerasnip models. Full conformal inference is provided for completeness and for situations where the extra statistical rigour is required and training cost is acceptable.

```{r full-conformal}
# Use a small subset to keep runtime reasonable in this vignette
data_small <- data[1:100, ]
new_obs_small <- data[101:106, ]

fit_small <- fit(wflow, data = data_small)

conformal_full <- int_conformal_full(
  fit_small,
  train_data = data_small,
  control = control_conformal_full(
    method = "grid",
    trial_points = 20
  )
)
conformal_full

predict(conformal_full, new_data = new_obs_small, level = 0.90)
```

**When to use this**: when you need the strongest possible coverage guarantees and are willing to pay the training cost. For neural networks, prefer `int_conformal_split` unless you have a specific reason to use full conformal inference.

---

## Comparison of the three methods

| Method | Model refits | Coverage guarantee | Recommended for kerasnip |
|---|---|---|---|
| `int_conformal_split` | 1 (on training data only) | Marginal | Yes, lowest cost |
| `int_conformal_cv` | 1 per fold + 1 final | Marginal | Yes, if already cross-validating |
| `int_conformal_full` | *(n_test × n_grid)* | Marginal | With caution, expensive |

All three methods provide **marginal coverage**: across many test observations, at least the requested fraction will have their true outcome within the interval. None of them provide **conditional coverage** (intervals that are exactly correct for every specific input value), but in practice they are well-calibrated for most regression problems.

---

## Cleanup

```{r cleanup}
remove_keras_spec("conf_mlp")
```
