---
title: "Saving, Reloading, and Sharing kerasnip Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Saving, Reloading, and Sharing kerasnip Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## Overview

Once you have fitted a kerasnip workflow, you will naturally want to save it —
whether to deploy it, share it with a collaborator, or simply reload it later
without re-training. This vignette explains why the default approaches fail and
what to do instead.

## Why plain `saveRDS()` fails

Saving a fitted workflow with `saveRDS()` and loading it in a new R session will
break prediction. There are two distinct root causes.

### Root cause 1: parsnip registrations are session-scoped

`create_keras_sequential_spec()` and `create_keras_functional_spec()` work by
making side-effect calls to parsnip's internal API (`set_new_model()`,
`set_model_mode()`, `set_model_engine()`, `set_fit()`, `set_pred()`, etc.).
These registrations live in parsnip's internal environment, which is **never
serialized**. After `library(kerasnip)` in a fresh R session, parsnip has no
knowledge of your custom model type.

When you call `predict()` on the loaded workflow, parsnip tries to look up the
encoding for your model type and fails:

```
Error in `get_encoding()`:
! Model "basic_mlp" has not been registered.
```

### Root cause 2: Keras Python objects cannot survive `saveRDS()`

The fitted Keras model inside your workflow stores references to Python objects.
These references are process-specific and are lost when R serializes the object
with `saveRDS()`. Loading the result in a new process leaves the Keras model in
a broken state, even before you try to predict.

## The recommended workflow

The `bundle` package (from the tidymodels ecosystem) solves root cause 2 by
using Keras's own native serialization to safely capture the Python model.
`kerasnip`'s `reregister_keras_spec()` function solves root cause 1 by
replaying the parsnip registration from metadata embedded in the spec object.
Together they give you a complete, portable save/load workflow.

### Setup

```{r setup}
library(kerasnip)
library(keras3)
library(workflows)
library(bundle)
```

### Session 1 — define, fit, and save

```{r session-1}
# Define your layer blocks as usual
input_block <- function(model, input_shape) {
    keras_model_sequential(input_shape = input_shape)
}
hidden_block <- function(model, units = 64) {
    model |> layer_dense(units = units, activation = "relu")
}
output_block <- function(model) {
    model |> layer_dense(units = 1)
}

# Create the spec
create_keras_sequential_spec(
    model_name = "basic_mlp",
    layer_blocks = list(
        input  = input_block,
        hidden = hidden_block,
        output = output_block
    ),
    mode = "regression"
)

# Build and fit a workflow
wf <- workflow() |>
    add_model(basic_mlp(
        fit_epochs = 10,
        compile_loss = "mean_squared_error"
    )) |>
    add_formula(mpg ~ .)

fit_wf <- fit(wf, data = mtcars)

# Bundle handles Keras native serialization; saveRDS handles the R object
bundled <- bundle(fit_wf)
saveRDS(bundled, file = "my_model.rds")
```

### Session 2 — load and predict

```{r session-2}
library(kerasnip)
library(bundle)

# 1. Load the bundled object
bundled <- readRDS("my_model.rds")

# 2. Unbundle — bundle restores the Keras Python model
fit_wf <- unbundle(bundled)

# 3. Re-register — kerasnip restores the parsnip registration
reregister_keras_spec(fit_wf)

# 4. Predict as usual
predict(fit_wf, new_data = head(mtcars))
```

## Lightweight alternative: skip `bundle` when weights aren't needed

If you only need to share the *specification* (for example, to restore the
parsnip registration and re-create the constructor function, but you intend to
re-fit the model from scratch), you can skip `bundle` entirely and use plain
`saveRDS()` combined with `reregister_keras_spec()`.

```{r plain-saverds}
# Session 1 — save only the fitted workflow (weights will be broken,
# but the spec metadata is sufficient for re-registration)
saveRDS(fit_wf, file = "my_spec.rds")

# Session 2
library(kerasnip)

fit_wf <- readRDS("my_spec.rds")
reregister_keras_spec(fit_wf) # restores parsnip registration + basic_mlp()

# You can now re-fit:
refit_wf <- fit(
    workflow() |>
        add_model(basic_mlp(fit_epochs = 10, compile_loss = "mse")) |>
        add_formula(mpg ~ .),
    data = mtcars
)
predict(refit_wf, new_data = head(mtcars))
```

Note that in this path, calling `predict()` directly on the loaded `fit_wf`
will still fail because the Keras weights are lost — `bundle` is required if you
need predictions from the original fitted model.

## What `reregister_keras_spec()` does

`reregister_keras_spec()` reads metadata that kerasnip (>= 0.1.1) automatically
embeds in every spec object at creation time — specifically the `layer_blocks`
list and whether the Functional or Sequential API was used. It then replays the
full parsnip registration and also re-creates the named constructor function
(e.g. `basic_mlp()`) in your calling environment.

It accepts either a spec object or a fitted workflow:

```{r reregister-usage}
# From a workflow (most common after loading)
reregister_keras_spec(fit_wf)

# From a bare spec
spec <- extract_spec_parsnip(fit_wf)
reregister_keras_spec(spec)
```

## Backward compatibility

The embedded metadata was introduced in kerasnip 0.1.1. Specs created with
older versions do not carry it. Calling `reregister_keras_spec()` on such a spec
will produce a clear error message directing you to call
`create_keras_sequential_spec()` or `create_keras_functional_spec()` manually.

## What to share with collaborators

To allow a collaborator to load your model and predict without re-training,
share the output of `bundle()` saved as an `.rds` file. They will need:

- `library(kerasnip)` — to load the package and get `reregister_keras_spec()`
- `library(bundle)` — to call `unbundle()`
- Then the three-line load sequence from Session 2 above
