---
title: "Saving and Reloading Fitted Kerasnip Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Saving and Reloading Fitted Kerasnip Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("keras")
)
# Suppress verbose Keras output for the vignette
options(keras.fit_verbose = 0)
set.seed(123)
```

## Overview

Keras models are backed by Python objects managed by TensorFlow/JAX. These objects live in the current R session and are represented as external pointers (`externalptr`) that become invalid as soon as the session ends, or even within the same session after `saveRDS()` / `readRDS()`.

`kerasnip` handles this transparently so that fitted workflows can be saved, reloaded, and used for prediction without any manual restoration steps.

## Quick workflow example

Before discussing the details, here is the full persistence workflow:

```{r load-packages, cache=TRUE}
library(kerasnip)
library(tidymodels)
library(keras3)

# 1. Define Layer Blocks (Required by kerasnip)
# The first block must initialize the sequential model
input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}

# Hidden layer block
dense_block <- function(model, units = 32) {
  model |> layer_dense(units = units, activation = "relu")
}

# Output layer block (units = 1 for regression)
output_block <- function(model, num_classes) {
  model |> layer_dense(units = 1)
}

# 2. Generate the parsnip specification
create_keras_sequential_spec(
  model_name = "my_mlp",
  layer_blocks = list(
    input = input_block,
    hidden = dense_block,
    output = output_block
  ),
  mode = "regression"
)

# 3. Use the newly created 'my_mlp' function
mod_spec <- my_mlp(fit_epochs = 10) |> 
  set_engine("keras")

# 4. Standard tidymodels workflow
rec_spec <- recipe(mpg ~ ., data = mtcars) |> 
  step_normalize(all_predictors())

fit_wf <- workflow() |> 
  add_recipe(rec_spec) |> 
  add_model(mod_spec) |> 
  fit(data = mtcars)

# Predict
new_data <- mtcars[1:3, ]
predict(fit_wf, new_data)
```

The first call to predict() detects that the Python pointer is invalid and restores the model from the stored bytes automatically.

## What kerasnip does behind the scenes

`kerasnip` handles persistence automatically:

- At **fit time**, the Keras model is serialized to a raw byte vector (`.keras`format) and stored alongside the parsnip `model_fit` object.
- At **predict time**, if the Python pointer is detected as invalid, `predict()` automatically restores the model from those bytes before dispatching.
- The parsnip model specification is also re-registered if it is missing from the session (e.g. after a fresh start).

This means you can use the persistence strategy that best suits your workflow without any extra boilerplate.

## Strategy 1: Plain `saveRDS()` / `readRDS()`

For most use cases: sharing a model file with a colleague, caching a fit between R sessions, or checkpointing during development; plain RDS is the simplest approach.

```{r rds-workflow}
library(kerasnip)
library(workflows)
library(parsnip)
library(recipes)

# --- Save ---
saveRDS(fit_wf, "my_model.rds")

# --- Reload in the same or a new R session ---
library(kerasnip)
fit_wf <- readRDS("my_model.rds")

# predict() restores the Keras model from bytes automatically
predictions <- predict(fit_wf, new_data = new_data)
predictions
```

There is nothing special to do after `readRDS()`. The first call to `predict()` detects the invalid pointer, restores the model from the stored bytes, and then proceeds normally.

## Strategy 2: `bundle` / `unbundle`

The [`bundle`](https://rstudio.github.io/bundle/) package provides a standardized serialization interface used by `vetiver`, `plumber`, and other MLOps tools. It is the right choice when:

- You are deploying a model to a `vetiver` API or a Docker container.
- You want a self-contained, version-controlled artifact that does not rely on any R session state.
- You are sharing a model across machines with different R library paths.

```{r bundle-workflow}
library(kerasnip)
library(bundle)
library(workflows)

# --- Save ---
bundled <- bundle(fit_wf)
saveRDS(bundled, "my_model_bundle.rds")

# --- Reload in any R session ---
library(kerasnip)
library(bundle)
bundled <- readRDS("my_model_bundle.rds")
fit_wf <- unbundle(bundled)
predictions <- predict(fit_wf, new_data = new_data)
predictions
```

## Comparison

| | `saveRDS` / `readRDS` | `bundle` / `unbundle` |
|---|---|---|
| Works across sessions | ✅ | ✅ |
| Works across machines | ✅ (same R library) | ✅ |
| `vetiver` / Docker compatible | ❌ | ✅ |
| Extra dependency needed | ❌ | `bundle` package |
| Code complexity | Minimal | Minimal |

## What happens under the hood

When `kerasnip` fits a model, the generic fit function calls `keras_model_to_bytes()`, which writes the model to a temporary `.keras` file using `keras3::save_model()` and reads the bytes back into R:

```{r internals, eval = FALSE}
# Simplified version of what happens inside generic_sequential_fit()
keras_bytes <- keras_model_to_bytes(model)
# keras_bytes is a raw vector stored in object$fit$keras_bytes
```

When `predict()` is called on a reloaded object, `predict.kerasnip_model_fit()` runs:

```{r restore-logic, eval = FALSE}
# Simplified version of predict.kerasnip_model_fit()
if (!is.null(object$fit$keras_bytes)) {
  is_valid <- tryCatch(
    {
      reticulate::py_validate_xptr(object$fit$fit)
      TRUE
    },
    error = function(e) FALSE
  )
  if (!is_valid) {
    object$fit$fit <- keras_model_from_bytes(object$fit$keras_bytes)
  }
}
```

If `keras_model_to_bytes()` fails (e.g. if the model was compiled with a non-serialisable custom object), a warning is issued at fit time and `keras_bytes` is set to `NULL`. In that case, `predict()` after reload will fail with a clear error.
