---
title: The Sequential Model with kerasnip
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Sequential Model with kerasnip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette provides a comprehensive guide to using `kerasnip` to define sequential Keras models within the `tidymodels` ecosystem. `kerasnip` bridges the gap between the imperative, layer-by-layer construction of Keras models and the declarative, specification-based approach of `tidymodels`.

Here, we will focus on `create_keras_sequential_spec()`, which is ideal for models where layers form a plain stack, with each layer having exactly one input tensor and one output tensor.

## Setup

We'll start by loading the necessary packages:

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Suppress verbose Keras output for the vignette
options(keras.fit_verbose = 0)
set.seed(123)
```

```{r load-packages}
library(kerasnip)
library(tidymodels)
library(keras3)
```

## When to use `create_keras_sequential_spec()`

A `Sequential` model in Keras is appropriate for a plain stack of layers where each layer has exactly one input tensor and one output tensor. `kerasnip`'s `create_keras_sequential_spec()` function is designed to define such models in a `tidymodels`-compatible way.

Instead of building the model layer-by-layer imperatively, you define a named, ordered list of R functions called `layer_blocks`. Each `layer_block` function takes a Keras model object as its first argument and returns the modified model. `kerasnip` then uses these blocks to construct the full Keras Sequential model.

For models with more complex, non-linear topologies (e.g., multiple inputs/outputs, residual connections, or multi-branch models), you should use `create_keras_functional_spec()`.

## Creating a `kerasnip` Sequential Model Specification

Let's define a simple sequential model with three dense layers.

First, we define our `layer_blocks`:

```{r define-simple-blocks}
# The first block must initialize the model. `input_shape` is passed automatically.
input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}

# A reusable block for hidden layers. `units` will become a tunable parameter.
hidden_block <- function(model, units = 32, activation = "relu") {
  model |> layer_dense(units = units, activation = activation)
}

# The output block. `num_classes` is passed automatically for classification.
output_block <- function(model, num_classes, activation = "softmax") {
  model |> layer_dense(units = num_classes, activation = activation)
}
```

Now, we use `create_keras_sequential_spec()` to generate our `parsnip` model specification function. We'll name our model `my_simple_mlp`.

```{r create-simple-spec}
create_keras_sequential_spec(
  model_name = "my_simple_mlp",
  layer_blocks = list(
    input = input_block,
    hidden_1 = hidden_block,
    hidden_2 = hidden_block,
    output = output_block
  ),
  mode = "classification"
)
```

## A common debugging workflow: `compile_keras_grid()`

In the original Keras guide, a common workflow is to incrementally add layers and call `summary()` to inspect the architecture. With `kerasnip`, the model is defined declaratively, so we can't inspect it layer-by-layer in the same way.

However, `kerasnip` provides a powerful equivalent: `compile_keras_grid()`. This function checks if your `layer_blocks` define a valid Keras model and returns the compiled model structure, all without running a full training cycle. This is perfect for debugging your architecture.

Let's see this in action with a CNN architecture:

```{r compile-grid-debug}
# Define CNN layer blocks
cnn_input_block <- function(model, input_shape) {
  keras_model_sequential(input_shape = input_shape)
}
cnn_conv_block <- function(model, filters = 32, kernel_size = 3, activation = "relu") {
  model |> layer_conv_2d(filters = filters, kernel_size = kernel_size, activation = activation)
}
cnn_pool_block <- function(model, pool_size = 2) {
  model |> layer_max_pooling_2d(pool_size = pool_size)
}
cnn_flatten_block <- function(model) {
  model |> layer_flatten()
}
cnn_output_block <- function(model, num_classes, activation = "softmax") {
  model |> layer_dense(units = num_classes, activation = activation)
}

# Create the kerasnip spec function
create_keras_sequential_spec(
  model_name = "my_cnn",
  layer_blocks = list(
    input = cnn_input_block,
    conv1 = cnn_conv_block,
    pool1 = cnn_pool_block,
    flatten = cnn_flatten_block,
    output = cnn_output_block
  ),
  mode = "classification"
)

# Create a spec instance for a 28x28x1 image
cnn_spec <- my_cnn(
  conv1_filters = 32, conv1_kernel_size = 5,
  compile_loss = "categorical_crossentropy",
  compile_optimizer = "adam"
)

# Prepare dummy data with the correct shape.
# We create a list of 28x28x1 arrays.
x_dummy_list <- lapply(1:10, function(i) array(runif(28*28*1), dim = c(28, 28, 1)))
x_dummy_df <- tibble::tibble(x = x_dummy_list)
y_dummy <- factor(sample(0:9, 10, replace = TRUE), levels = 0:9)
y_dummy_df <- tibble::tibble(y = y_dummy)


# Use compile_keras_grid to get the model summary
compilation_results <- compile_keras_grid(
  spec = cnn_spec,
  grid = tibble::tibble(), 
  x = x_dummy_df,
  y = y_dummy_df
)

# Print the summary
summary(compilation_results$compiled_model[[1]])
```

<!-- ## What to do once you have a model -->
<!-- -->
<!-- Once your model architecture is ready, you will want to: -->
<!-- -->
<!-- TODO: Create vignettes for the following guides and uncomment these links: -->
<!-- -->
<!-- - [Train your model, evaluate it, and run inference](https://keras3.posit.co/articles/training_and_evaluation.html) -->
<!-- - [Save your model to disk and restore it](https://keras3.posit.co/articles/serialization_and_saving.html) -->

## Feature Extraction with a Sequential Model

Once a Sequential model has been built, it behaves like a Functional API model. This means that every layer has an input and output attribute. In `kerasnip`, we can get access to this underlying model structure using `compile_keras_grid()`.

This allows us to create a new model that outputs the values of the intermediate layers.

```{r feature-extraction}
# We can reuse the compilation results from the previous chunk
keras_model_obj <- compilation_results$compiled_model[[1]]

# Create a new Keras model for feature extraction
feature_extractor <- keras_model(
  inputs = keras_model_obj$inputs,
  outputs = lapply(keras_model_obj$layers, function(x) x$output)
)

# Call the feature extractor on a dummy input tensor
x_tensor <- op_ones(c(1, 28, 28, 1))
features <- feature_extractor(x_tensor)

# Print the shapes of the extracted feature maps
lapply(features, dim)
```

## Transfer Learning with a Sequential Model

<!-- TODO: Create a vignette for the transfer learning guide and uncomment this link: -->
<!-- If you aren’t familiar with it, make sure to read our [guide to transfer learning](https://keras3.posit.co/articles/transfer_learning.html). -->

Transfer learning consists of freezing the bottom layers in a model and only training the top layers. A common blueprint is to use a Sequential model to stack a pre-trained model and some freshly initialized classification layers.

`kerasnip` supports this by allowing a `layer_block` to contain a pre-trained model.

```{r transfer-learning}
# Define a block that incorporates a pre-trained base
# This block creates a new sequential model and adds the pre-trained,
# frozen base model as its first layer.
pretrained_base_block <- function(model, input_shape) {
  base_model <- application_xception(
    weights = "imagenet",
    include_top = FALSE,
    pooling = "avg",
    input_shape = input_shape
  )
  # Freeze the weights of the pre-trained base
  freeze_weights(base_model)
  
  # The block must return a sequential model
  keras_model_sequential(input_shape = input_shape) |>
    base_model()
}

# Define a new classification head. This block will be appended to the
# sequential model returned by the previous block.
classification_head_block <- function(model, num_classes) {
  model |>
    layer_dense(units = 1000, activation = "relu") |> 
    layer_dense(units = num_classes, activation = "softmax")
}

# Create a new kerasnip spec with the pre-trained base and new head
create_keras_sequential_spec(
  model_name = "transfer_cnn",
  layer_blocks = list(
    base = pretrained_base_block,
    head = classification_head_block
  ),
  mode = "classification"
)

# Create a spec instance
transfer_spec <- transfer_cnn(
  compile_loss = "categorical_crossentropy",
  compile_optimizer = "adam"
)

# Prepare dummy data for a 224x224x3 image
x_dummy_tl_list <- lapply(1:10, function(i) array(runif(224*224*3), dim = c(224, 224, 3)))
x_dummy_tl_df <- tibble::tibble(x = x_dummy_tl_list)
y_dummy_tl <- factor(sample(0:9, 10, replace = TRUE), levels = 0:9)
y_dummy_tl_df <- tibble::tibble(y = y_dummy_tl)


# Use compile_keras_grid to inspect the model and trainable parameters
compilation_results_tl <- compile_keras_grid(
  spec = transfer_spec,
  grid = tibble::tibble(),
  x = x_dummy_tl_df,
  y = y_dummy_tl_df
)

# Print the summary to verify that the base model's parameters are non-trainable
summary(compilation_results_tl$compiled_model[[1]])
```

<!-- To find out more about building models in Keras, see: -->

<!-- TODO: Create vignettes for the following guides and uncomment these links: -->
<!-- -->
<!-- - [The Functional API](https://keras3.posit.co/articles/functional_api.html) -->
<!-- - [Making new Layers & Models via subclassing](https://keras3.posit.co/articles/making_new_layers_and_models_via_subclassing.html) -->